\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-06-29 analysis/paper/paper.Rmd

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(tidyverse)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

@@

@@

```{r erspc-cox-cif, eval = eval_cs1, echo = echo_plot_code}
new_data <- data.frame(
  ScrArm = c("Control group", "Screening group"),
  ignore = 99
)

plot(survfit(cox_model, newdata = new_data),
  xlab = "Years since Randomization",
  ylab = "Cumulative Incidence (%)",
  fun = "event",
  xlim = c(0, 15), conf.int = FALSE, col = c("red", "blue"),
  main = sprintf(
    "Estimated Cumulative Incidence (risk) of Death from Prostate 
                    Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
    exp(coef(cox_model)),
    exp(confint(cox_model))[1],
    exp(confint(cox_model))[2]
  ),
  ylim = c(0, 2), yscale = 100
)
legend("topleft",
  legend = c("Control group", "Screening group"),
  col = c("red", "blue"),
  lty = c(1, 1),
  bg = "gray90"
)
```

@@

```{r erspc-casebase-exponential, eval = eval_cs1}
casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm,
  data = ERSPC,
  ratio = 100
)

summary(casebase_exponential)
exp(coef(casebase_exponential)[2])
exp(confint(casebase_exponential)[2, ])
```

@@

summary(casebase_time)
exp(coef(casebase_time))
exp(confint(casebase_time))
```

@@

@@

>>>>

---
documentclass: jss
author:
  - name: Sahir Bhatnagar *
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon *
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. * joint co-authors
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
output: 
  rticles::jss_article:
    number_sections: TRUE     #added argument option 
    citation_package: "natbib"  #All my citations use biblatex, not natbib. 
biblio-style: jss      #Listed to use in JSS Instructions for Authors, but not in template by default. 
bibliography: references.bib  #Also not included in template by default. 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = TRUE,
  comment = "#>",
  fig.path = "../figures/"
)
options(kableExtra.latex.load_packages = FALSE)
library(casebase)
library(colorspace)
library(cowplot)
library(data.table)
library(glmnet)
library(kableExtra)
library(magrittr)
library(pracma)
library(flexsurv) # for fitting gompertz AFT
library(prodlim)
library(riskRegression)
library(splines)
library(survival)
library(tibble)
library(tidyverse)
options(digits = 2, scipen = 999)
# set the seed for reproducible output
set.seed(1234)

@@

The case-base sampling approach described in Section \ref{theory} can be visualized in the form of a population time plot. They are extremely informative graphical displays of survival data and should be one of the first steps in an exploratory data analysis. We facilitate this task in the \pkg{casebase} package using the \code{popTime} function and \code{plot} method as follows:

The splitting of these tasks gives the user the flexibility to create population-time plots with the graphics system of their choice using the object created in Step 1, while the object created in Step 2 allows one to leverage the \pkg{ggplot2} family of functions for further customization. 
1. The \code{casebase::popTime} function takes as input the original dataset along with the column names corresponding to the timescale, the event status and an exposure group of interest (optional). This will create an object of class \code{popTime} (or \code{popTimeExposure} if you specify a value for the exposure argument).  
2. The corresponding \code{plot} methods for the object created in Step 1 can be called to create the population time plot with several options for customizing the aesthetics.

To illustrate these functions we will use the European Randomized Study of Prostate Cancer Screening data which is available through the \pkg{casebase} package. The results of this study were published by [@schroder2009screening], and the dataset itself was obtained using the approach described in [@liu2014recovering].

We first create the necessary dataset for producing the population time plots using the \code{casebase::popTime} function. In this example we stratify by the treatment group. Note that the resulting object stores the exposure variable as an attribute:

We first pass this dataset to the \code{popTime} function. If you do not specify the time and event arguments, this function will try to guess them using regular expressions. We then generate the plot by using the corresponding \code{plot} method:
```{r plot-erspc-data, eval = eval_cs1}
data("ERSPC")
ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
                       levels = c(0,1), 
                       labels = c("Control group", "Screening group"))
pt_object <- casebase::popTime(ERSPC, 
                               time = "Follow.Up.Time",
                               event = "DeadOfPrCa",
                               exposure = "ScrArm")
attr(pt_object, "exposure")
```

<!--The \code{pt_object} is a \code{data.table} containing the original dataset augmented with-->
We then pass this object to the \code{plot} function which leverages the \pkg{ggplot2} package [@ggplot2] to build population time plots. It builds the plot by sequentially adding layers, starting with a layer for the area representing the population time followed by points to show the casebase sampling mechanism. This function gives user the flexibility to add any combination of the case.series, base.series and competing events. The case series and competing events are sampled at random vertically on the plot in order to visualise the incidence density using the popTime function. That is, imagine we draw a vertical line at a specific event time. We then plot the point at a randomly sampled y-coordinate along this vertical line. This is done to avoid having all points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis). By randomly distributing them, we can get a better sense of the inicidence density. The base series is sampled horizontally on the plot using the sampleCaseBase function.

One benefit from these plots is that it allows you to see the incidence density. This can be seen from the distribution of the red dots in the above plot. We can see that more events are observed later on in time. Therefore a constant hazard model would not be appropriate in this instance as it would overestimate the cumulative incidence earlier on in time, and underestimate it later on.

The unique _step shape_ of the population time plot is due to the randomization of the Finnish cohorts which were carried out on January 1 of each of the 4 years 1996 to 1999. This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up.

It is important to note that the red points are random distributed across the grey area for each time of event. That is, imagine we draw a vertical line at a specific event time. We then plot the red point at a randomly sampled y-coordinate along this vertical line. This is done to avoid having all red points along the upper edge of the plot (because the subjects with the least amount of observation time are plotted at the top of the y-axis). By randomly distributing them, we can get a better sense of the inicidence density.

The randomization of the Finnish cohorts were carried out on January 1 of each of the 4 years 1996 to 1999. This, coupled with the uniform December 31 2006 censoring date, lead to large numbers of men with exactly 11, 10, 9 or 8 years of follow-up. Tracked backwards in time (i.e. from right to left), the population-time plot shows the recruitment pattern from its beginning in 1991, and in particular the Jan 1 entries in successive years. Tracked forwards in time (i.e. from left to right), the plot for the first three years shows attrition due entirely to death (mainly from other causes). Since the Swedish and Belgian centres were the last to close their recruitment - in December 2003 - the minimum potential follow-up is three years. Tracked further forwards in time (i.e. after year 3) the attrition is
We can also create exposure stratified plots by specifying the \code{exposure} argument in the \code{popTime} function:

```{r plot-stratified-erspc-data, eval = eval_cs1}
pt_object_strat <- casebase::popTime(ERSPC, 
                                     event = "DeadOfPrCa", 
                                     exposure = "ScrArm")
# class(pt_object_strat)
# pt_object
plot(pt_object_strat)
# dev.off()
# library(magrittr)
# pt <- plot(pt_object_strat)
# pt %>% class

# library(ggplot2)
# pt + theme_minimal()

```

```{r plot-stratified-erspc-data-side, eval = eval_cs1}
plot(pt_object_strat, ncol = 2)
```

For a more detailed description of how to modify the aesthetics of a population time plot, please refer to the [Customizing Population Time Plots](http://sahirbhatnagar.com/casebase/articles/customizingpopTime.html) vignette.

@@

```{r erspc-cox-cif, eval = eval_cs1, echo = echo_plot_code, fig.width=8, fig.height=6, fig.cap="Cumulative incidence functions for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression. (A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution. (B) Linear function of time which is equivalent to a Gompertz distribution. (C) The natural logarithm of time which is equivalent to a Weibull distribution. (D) Cubic b-spline expansion of time."}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
myCols <- cbPalette[c(4,7)]
new_data <- data.frame(
  ScrArm = c("Control group", "Screening group"),
  ignore = 99
)

fmla <- list(exponential = as.formula(DeadOfPrCa ~ ScrArm),
             gompertz = as.formula(DeadOfPrCa ~ Follow.Up.Time + ScrArm),
             weibull = as.formula(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm),
             splines = as.formula(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm))

fits <- lapply(fmla, function(i) {
  casebase::fitSmoothHazard(i, data = ERSPC, ratio = 100)
})

risks <- lapply(fits, function(i) {
  casebase::absoluteRisk(object = i, time = seq(0,15,0.1), newdata = new_data)
})

aft_fits <- lapply(list("exp","gompertz","weibull"), function(i) 
  flexsurv::flexsurvreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
                           data = ERSPC, dist = i)
  )

names(aft_fits) <- list("exponential","gompertz","weibull")

# dev.off()
layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
     #                exp(coef(cox_model)), 
     #                exp(confint(cox_model))[1], 
     #                exp(confint(cox_model))[2]), 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Casebase)",
                  "Screening (Cox)", "Screening (Casebase)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(A)", cex = 1.5)

par(mar = c(4.1, 4.1, 0, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "", xlab = "",
     fun = "event",
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$weibull[,1], risks$weibull[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$weibull[,1], risks$weibull[,3], col = myCols[2], lty = 2, lwd = 2)  
text(x = 15*0.97, y = 0.05/100, labels = "(C)", cex = 1.5)
mtext("Cumulative incidence (%)", side = 2, line = -1.5, outer = TRUE)

par(mar = c(0, 0, 4.1, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     xaxt = 'n', yaxt = 'n',
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$gompertz[,1], risks$gompertz[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$gompertz[,1], risks$gompertz[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(B)", cex = 1.5)

par(mar = c(4.1, 0, 0, 3.9))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",xlab = "",
     fun = "event",
     yaxt = 'n',
     xlim = c(0,15), conf.int = FALSE, col = myCols, 
     lwd = 2,
     yscale = 100)
lines(risks$splines[,1], risks$splines[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$splines[,1], risks$splines[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = 15*0.97, y = 0.05/100, labels = "(D)", cex = 1.5)
mtext("Years since randomization", side = 1, line = -1.5, outer = TRUE)
```

@@

## Equivalence between casebase and AFT

```{r erspc-casebase-exponential, eval = eval_cs1}
# exponential -------------------------------------------------------------

casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
                                                  data = ERSPC, 
                                                  ratio = 100)
aft_exponential <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
                           data = ERSPC, dist = "exp")

-coef(aft_exponential)
coef(casebase_exponential)
summary(casebase_exponential)
exp(coef(casebase_exponential)[2])
exp(confint(casebase_exponential)[2, ])

# weibull -----------------------------------------------------------------

casebase_weibull <- casebase::fitSmoothHazard(DeadOfPrCa ~ log(Follow.Up.Time) + ScrArm, 
                                           data = ERSPC, 
                                           ratio = 100)
aft_weibull <- survreg(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
                       data = ERSPC, dist = "weibull")

-coef(aft_weibull)/aft_weibull$scale
coef(casebase_weibull)
```
We can then use the \code{absoluteRisk} function to get an estimate of the cumulative incidence curve for a specific covariate profile. In the plot below, we overlay the estimated CIF from the exponential model on the Cox model CIF:

@@

@@

```{r erspc-95ci, echo = FALSE, eval = eval_cs1}
# CI_cox <- c(exp(coef(cox_model)), exp(confint(cox_model)))
# CI_exp <- c(exp(coef(fits$exponential))["ScrArmScreening group"], exp(confint(fits$exponential))["ScrArmScreening group",])
# CI_gompertz <- c(exp(coef(fits$gompertz))["ScrArmScreening group"], exp(confint(fits$gompertz))["ScrArmScreening group",])
# CI_weibull <- c(exp(coef(fits$weibull))["ScrArmScreening group"], exp(confint(fits$weibull))["ScrArmScreening group",])
# CI_splines <- c(exp(coef(fits$splines))["ScrArmScreening group"], exp(confint(fits$splines))["ScrArmScreening group",])

# function to exponentiate HR and CI and return a HR (lower, Upper) character
exp_est_ci <- function(model = c("exponential","gompertz","weibull","splines"),
                       fit, 
                       covariate = "ScrArmScreening group", 
                       type = c("casebase", "survreg")) {

  type <- match.arg(type)
  covariate <- match.arg(covariate)

  if (type == "casebase"){

    cis <- confint(fit[[model]])

    sprintf("%0.2f (%0.2f, %0.2f)",
            exp(coef(fit[[model]]))[[covariate]],
            exp(cis)[covariate,1],
            exp(cis)[covariate,2])
  } else {

    if (model == "weibull") {

      # The maximised log-likelihoods are the same between survreg and flexsurvreg, 
      # however the parameterisation is different: the first coefficient (Intercept) 
      # reported by survreg is log(mu), and survreg's "scale" is
      # dweibull’s (thus flexsurvreg)'s 1 / shape
      res <- -fit[[model]]$res[covariate,] / (1 / fit[[model]]$res["shape",])

      sprintf("%0.2f (%0.2f, %0.2f)",
              exp(res[["est"]]),
              exp(res[["U95%"]]),
              exp(res[["L95%"]]))

    } else if (model == "splines") {

      c("--")

    } else {

      res <- fit[[model]]$res[covariate,]

      sprintf("%0.2f (%0.2f, %0.2f)",
              exp(res[["est"]]),
              exp(res[["L95%"]]),
              exp(res[["U95%"]]))
    }
  }
}

casebase_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
                         exp_est_ci, 
                         fit = fits,
                         covariate = "ScrArmScreening group", 
                         type = "casebase")

survreg_HR_CI <- lapply(list("exponential","gompertz","weibull","splines"), 
                         exp_est_ci, 
                         fit = aft_fits,
                         covariate = "ScrArmScreening group", 
                         type = "survreg")

pt <- cbind(list("Exponential","Gompertz","Weibull","Splines"),
            do.call(rbind, casebase_HR_CI), do.call(rbind, survreg_HR_CI))

# table_ci <- rbind(CI_cox, CI_exp, CI_gompertz, CI_weibull, CI_splines)
# colnames(table_ci) <- c("HR", "LB", "UB")
# 
# data.frame(Model = c("Cox", "Exponential", "Gompertz", "Weibull", "Splines")) %>% 
#   bind_cols(as.data.frame(table_ci)) %>% 
#   mutate(CI = paste0("(", round(LB, 2), ", ",
#                      round(UB, 2), ")"),
#          Width = UB - LB) %>% 
#   dplyr::select(-LB, -UB) %>% 
#   knitr::kable(format = "latex", booktabs = TRUE)  %>%
#   kable_styling()

@@

```{r, eval = TRUE}
## ---- print-sim-table ----

kable(pt, "latex", booktabs = T, align = c("l","c","c"),
      caption = c("Mean "),
      col.names = c("Model","casebase","survreg")) %>%
  kable_styling() %>%
  # add_header_above(c(" "," ","No overlap" = 2, "All causal SNPs\nin kinship" = 2, "No overlap" = 2, "All causal SNPs\nin kinship" = 2)) %>%
  # add_header_above(c(" "," ","Null model" = 4, "1% Causal SNPs" = 4)) %>%
  column_spec(1, bold=T) %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
  footnote(general = "Median (Inter-quartile range) is given for Model Size.",
           general_title = "Note:") #%>%
  # footnote(general = "Model Size is the number of selected variables using the high-dimensional BIC for \texttt{ggmix} and 10-fold cross validation for \texttt{lasso} and \texttt{twostep}.")
  #          # number = c("", "Footnote 2; "),
  #          # alphabet = c("Footnote A; ", "Footnote B; "),
  #          # symbol = c("Footnote Symbol 1; ", "Footnote Symbol 2"),
  #          general_title = "Note: ", number_title = "Type I: ",
  #          alphabet_title = "Type II: ", symbol_title = "Type III: ",
  #          footnote_as_chunk = T, title_format = c("italic", "underline")
  # )
```

### 2020-06-29 analysis/paper/paper.tex

@@

@@

\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}

@@

\begin{CodeChunk}
\begin{table}[H]
\centering
\begin{tabular}{lrlr}
\toprule
Model & HR & CI & Width\\
\midrule
Cox & 0.80 & (0.67, 0.95) & 0.28\\
Exponential & 0.81 & (0.68, 0.96) & 0.28\\
Weibull & 0.79 & (0.67, 0.94) & 0.28\\
Splines & 0.81 & (0.68, 0.96) & 0.28\\
\bottomrule
\end{tabular}
\end{table}

>>>>

The case-base sampling approach described in Section \ref{theory} can be
visualized in the form of a population time plot. They are extremely
informative graphical displays of survival data and should be one of the
first steps in an exploratory data analysis. We facilitate this task in
the \pkg{casebase} package using the \code{popTime} function and
\code{plot} method as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The \code{casebase::popTime} function takes as input the original
  dataset along with the column names corresponding to the timescale,
  the event status and an exposure group of interest (optional). This
  will create an object of class \code{popTime} (or
  \code{popTimeExposure} if you specify a value for the exposure
  argument).\\
\item
  The corresponding \code{plot} methods for the object created in Step 1
  can be called to create the population time plot with several options
  for customizing the aesthetics.
\end{enumerate}

To illustrate these functions we will use the European Randomized Study
of Prostate Cancer Screening data which is available through the
\pkg{casebase} package. The results of this study were published by
\citep{schroder2009screening}, and the dataset itself was obtained using
the approach described in \citep{liu2014recovering}.

We first create the necessary dataset for producing the population time
plots using the \code{casebase::popTime} function. In this example we
stratify by the treatment group. Note that the resulting object stores
the exposure variable as an attribute:

\begin{CodeOutput}
#> [1] "ScrArm"
\end{CodeOutput}
\end{CodeChunk}

We then pass this object to the \code{plot} function which leverages the
\pkg{ggplot2} package \citep{ggplot2} to build population time plots. It
builds the plot by sequentially adding layers, starting with a layer for
the area representing the population time followed by points to show the
casebase sampling mechanism. This function gives user the flexibility to
add any combination of the case.series, base.series and competing
events. The case series and competing events are sampled at random
vertically on the plot in order to visualise the incidence density using
the popTime function. That is, imagine we draw a vertical line at a
specific event time. We then plot the point at a randomly sampled
y-coordinate along this vertical line. This is done to avoid having all
points along the upper edge of the plot (because the subjects with the
least amount of observation time are plotted at the top of the y-axis).
By randomly distributing them, we can get a better sense of the
inicidence density. The base series is sampled horizontally on the plot
using the sampleCaseBase function.

One benefit from these plots is that it allows you to see the incidence
density. This can be seen from the distribution of the red dots in the
above plot. We can see that more events are observed later on in time.
Therefore a constant hazard model would not be appropriate in this
instance as it would overestimate the cumulative incidence earlier on in
time, and underestimate it later on.

The unique \emph{step shape} of the population time plot is due to the
randomization of the Finnish cohorts which were carried out on January 1
of each of the 4 years 1996 to 1999. This, coupled with the uniform
December 31 2006 censoring date, lead to large numbers of men with
exactly 11, 10, 9 or 8 years of follow-up.

It is important to note that the red points are random distributed
across the grey area for each time of event. That is, imagine we draw a
vertical line at a specific event time. We then plot the red point at a
randomly sampled y-coordinate along this vertical line. This is done to
avoid having all red points along the upper edge of the plot (because
the subjects with the least amount of observation time are plotted at
the top of the y-axis). By randomly distributing them, we can get a
better sense of the inicidence density.

For a more detailed description of how to modify the aesthetics of a
population time plot, please refer to the
\href{http://sahirbhatnagar.com/casebase/articles/customizingpopTime.html}{Customizing
Population Time Plots} vignette.

@@

\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression. (A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution. (B) Linear function of time which is equivalent to a Gompertz distribution. (C) The natural logarithm of time which is equivalent to a Weibull distribution. (D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{CodeChunk}

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-2-1} \end{center}

@@

\caption{\label{tab:unnamed-chunk-4}Mean }
\centering
\begin{tabular}[t]{>{\bfseries}lcc}
\toprule
Model & casebase & survreg\\
\midrule
Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
Gompertz & 0.79 (0.66, 0.94) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
Splines & 0.80 (0.67, 0.95) & --\\
\bottomrule
\multicolumn{3}{l}{\textit{Note:}}\\
\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
\end{tabular}
\end{table}
